---
title: "Model bark thickness"
author: "Harriet, Eli, Michael"
date: "2026-06-15"
output:
  word_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

library(bayesplot)
library(dplyr, warn.conflicts = FALSE)
library(glue)
library(gratia)
library(mgcv)
library(sf)
library(tidybayes)

library(ggplot2)
theme_set( theme_bw() )

library(ggrepel)

# Package with function to extract random effect estimates from a GAM.
# This is not on CRAN but can be installed from github:
# remotes::install_github('m-clark/mixedup')
library(mixedup)


DATA_DIR <- here("data")
stopifnot(dir.exists(DATA_DIR))

MODEL_DIR <- here("models")
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

OUTPUT_DIR <- here("outputs")
if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)

GPKG <- file.path(DATA_DIR, "harriet_bark_dist.gpkg")
stopifnot(file.exists(GPKG))


# Function to use with dplyr::reframe() when drawing random effect estimates
fn_qdf <- function(x) {
  probs <- c(0.05, 0.25, 0.5, 0.75, 0.95)
  prob_names = sprintf("q%02d", probs*100)
  
  res <- data.frame(
    val = unname(quantile(x, probs = probs)),
    q = prob_names
  )
}


# Seed value to use with gratia for reproducible predictions between runs
SEED <- 42

# Flat to force refitting of all GAMs, even if cached models exist on disk
FORCE_REFIT <- FALSE

```


## Summary

Fit a generalized additive models to the bark thickness observations collected by Harriet and Eli. Separate models are fitted for each of the three broad bark types: smooth, stringy and rough.

The structure of the models was decided a-priori, i.e. we are not doing model selection and are not concerned if one or more terms do not appear to be influential. The goal is inference via model predictions.

Each model includes two random effect terms:
  - site-level: to account for the clustering of trees within survey sites;
  - species-level: to capture any systematic variation between species.
  
Note 2026-06-11: The original set of environmental predictors for the bark thickness models were:

  - landscape aridity
  - fire frequency
  - elevation
  - slope
  - broad lithology class
  
We subsequently decided to drop the following:
  - broad lithology, since manual inspection of the site values suggested it was too broad to be informative
  - elevation and slope, since these are effectively contained within the aridity index.


## Data

The data are stored as two non-spatial layers in the project geopackage. 

Tree-level observations...

```{r}

DAT_trees_thickness <- st_read(GPKG, "trees_thickness", quiet = TRUE)

str(DAT_trees_thickness)

```

Site-level predictors...

```{r}

DAT_predictors <- st_read(GPKG, "predictors_all", quiet = TRUE)

```


Sites...

```{r}

DAT_sites_thickness <- st_read(GPKG, "sites_thickness", quiet = TRUE)

```



Prepare a training data frame that includes all three broad bark types: smooth, stringy and rough.

```{r}

dat_all_barks <- DAT_trees_thickness %>%
  # Subset to records with thickness values for selected bark types
  filter(for_analysis, 
         !is.na(bark_thickness),
         bark_type_h %in% c("Smooth", "Stringy", "Rough")) %>%
  
  # Drop records for indet taxa (e.g. Eucalyptus 'unknown mallee')
  filter(!grepl(display_name, pattern = "unknown", ignore.case = TRUE)) %>%
  
  select(site_label, display_name, bark_type_h, bark_thickness, dbh) %>%

  mutate(f_species = factor(display_name)) %>%
  select(-display_name) %>%

  left_join(DAT_predictors, by = join_by(site_label)) %>%

  mutate(f_site = factor(site_label))

# Check data structure
str(dat_all_barks)

```

### Exploratory data summaries and graphs

Examine the number of survey sites in each combination of bark type and fire frequency.

```{r}

dat_all_barks %>%
  group_by(bark_type_h, fire_frequency) %>%
  summarize(nsites = n_distinct(f_site), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = bark_type_h, values_from = nsites, values_fill = 0)

```

For all bark types, there are very few sites in the lowest fire frequency classes (0, 1) or in the highest class (6) 

Look at the spatial distribution of the fire frequency classes...

```{r fig.width=10}

dat <- dat_all_barks %>%
  distinct(site_label, fire_frequency) %>%
  left_join(DAT_sites_thickness %>% select(site_label, geom), by = join_by(site_label)) %>%
  mutate(f_firefreq = factor(fire_frequency)) %>%
  st_as_sf() %>%
  bind_cols(st_coordinates(.))

ggplot(data = dat) +
  geom_sf(aes(colour = f_firefreq), size = 2, alpha = 0.4) +
  geom_text_repel(aes(x = X, y = Y, label = f_firefreq), max.overlaps = 50)

```


There is reasonable spatial mixing of the fire frequency classes other than the unburnt sites, which are infrequent and mostly in one small locality.


From trial modelling, we expect that the aridity index will be the most influential predictor on bark thickness. Here we explore the relationship between bark thickness and aridity within each fire frequency class. In particular, we are looking for any potential artefacts due to gaps in the sample coverage across space and/or across fire frequency.


```{r}

ggplot(data = dat_all_barks, aes(x = aridity, y = bark_thickness, group = bark_type_h, colour = bark_type_h)) +
  geom_point(alpha = 0.4) +
  geom_smooth(method = "gam", formula = y ~ s(x, bs="cr", k=3), se=FALSE) +
  facet_wrap(~fire_frequency, labeller = label_both)

```


The above graph shows:

  - There are few observations for the fire frequency classes 0 (unburnt), 1 and 6.
  
  - Unburnt (fire frequency = 0) sites display a positive trend for smooth and stringy bark thickness which is quite likely to be a sampling artefact.
  
  - There are GAM smooths with obvious problems for a couple of cases where data are particularly sparse, e.g. stringy bark thickness in fire frequency = 6.


### Define broader fire frequency classes

Based on the above data exploration, we lump fire frequency into two classes: less frequent (0 - 3 fires recorded) versus more frequency (4 - 6 fires recorded).

```{r}

dat_all_barks <- dat_all_barks %>%
  mutate(f_firefreq = factor(ifelse(fire_frequency <= 3, "Less frequent fire", "More frequent fire")))

```


Summary table for the lumped fire frequency classes across bark types:

```{r}

dat_all_barks %>%
  group_by(bark_type_h, f_firefreq) %>%
  summarize(nsites = n_distinct(f_site), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = bark_type_h, values_from = nsites, values_fill = 0)

```



## Model smooth bark thickness

```{r}

dat_smooth_bark <- dat_all_barks %>%
  filter(bark_type_h == "Smooth") %>%
  mutate(dbh_log = log(dbh))
  
glue("{nrow(dat_smooth_bark)} observations for {n_distinct(dat_smooth_bark$site_label)} sites")

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_smooth_bark_thickness.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_smooth_bark <- readRDS(outpath)
  
} else {
  m_smooth_bark <- gam(
    bark_thickness ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_smooth_bark,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_smooth_bark, outpath)
}

summary(m_smooth_bark)

```

Posterior predictive check. Compare the distribution of bark thickness values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_smooth_bark, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_smooth_bark$bark_thickness,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_smooth_bark, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_thick_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_smooth_bark, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_smooth_bark, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_thick_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```

### Counter-factual predictions

Derive prediction with DBH held at its median value, and the species and site random effects excluded.

```{r}

d_ <- dat_smooth_bark

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- dat_smooth_bark %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_smooth_bark, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_hdi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_smooth_bark,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_thick_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_smooth_bark

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_thick_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```


## Model stringy bark thickness

```{r}

dat_stringy_bark <- dat_all_barks %>%
  filter(bark_type_h == "Stringy") %>%
  mutate(dbh_log = log(dbh))
  
glue("{nrow(dat_stringy_bark)} observations for {n_distinct(dat_stringy_bark$site_label)} sites")

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_stringy_bark_thickness.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_stringy_bark <- readRDS(outpath)
  
} else {
  m_stringy_bark <- gam(
    bark_thickness ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_stringy_bark,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_stringy_bark, outpath)
}

summary(m_stringy_bark)

```

Posterior predictive check. Compare the distribution of bark thickness values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_stringy_bark, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_stringy_bark$bark_thickness,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_stringy_bark, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_thick_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_stringy_bark, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_stringy_bark, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_thick_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```


### Counter-factual predictions

Derive prediction with DBH held at its median value, and the species and site random effects excluded.

```{r}

d_ <- dat_stringy_bark

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- dat_stringy_bark %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_stringy_bark, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_hdi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_stringy_bark,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_thick_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_stringy_bark

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_thick_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```


## Model rough bark thickness

```{r}

dat_rough_bark <- dat_all_barks %>%
  filter(bark_type_h == "Rough") %>%
  mutate(dbh_log = log(dbh))

glue("{nrow(dat_rough_bark)} observations for {n_distinct(dat_rough_bark$site_label)} sites")

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_rough_bark_thickness.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_rough_bark <- readRDS(outpath)
  
} else {
  m_rough_bark <- gam(
    bark_thickness ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_rough_bark,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_rough_bark, outpath)
}

summary(m_rough_bark)

```


Posterior predictive check. Compare the distribution of bark thickness values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_rough_bark, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_rough_bark$bark_thickness,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_rough_bark, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_thick_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_rough_bark, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_rough_bark, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_thick_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```


### Counter-factual predictions

Derive prediction with DBH held at its median value, and the species and site random effects excluded.

```{r}

d_ <- dat_rough_bark

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- dat_rough_bark %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_rough_bark, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_hdi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_rough_bark,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_thick_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_rough_bark

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_thickness), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_thickness, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark thickness", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_thick_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```
