---
title: "Model bark density"
author: "Harriet, Eli and Michael"
date: "2026-06-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(here)

library(bayesplot)
library(dplyr, warn.conflicts = FALSE)
library(glue)
library(gratia)
library(mgcv)
library(sf)
library(tidybayes)

# Package with function to extract random effect estimates from a GAM.
# This is not on CRAN but can be installed from github:
# remotes::install_github('m-clark/mixedup')
library(mixedup)

library(ggplot2)
theme_set( theme_bw() )

library(ggrepel)


DATA_DIR <- here("data")
stopifnot(dir.exists(DATA_DIR))

MODEL_DIR <- here("models")
if (!dir.exists(MODEL_DIR)) dir.create(MODEL_DIR)

OUTPUT_DIR <- here("outputs")
if (!dir.exists(OUTPUT_DIR)) dir.create(OUTPUT_DIR)

GPKG <- file.path(DATA_DIR, "harriet_bark_dist.gpkg")
stopifnot(file.exists(GPKG))


# Function to use with dplyr::reframe() when drawing random effect estimates
fn_qdf <- function(x) {
  probs <- c(0.05, 0.25, 0.5, 0.75, 0.95)
  prob_names = sprintf("q%02d", probs*100)
  
  res <- data.frame(
    val = unname(quantile(x, probs = probs)),
    q = prob_names
  )
}


# Seed value to use with gratia for reproducible predictions between runs
SEED <- 42

# Flat to force refitting of all GAMs, even if cached models exist on disk
FORCE_REFIT <- FALSE

```


## Summary

Fit a generalized additive models to the bark density observations collected by Harriet and Eli. Separate models are fitted for each of the three broad bark types: smooth, stringy and rough.

The structure of the models was decided a-priori, i.e. we are not doing model selection and are not concerned if one or more terms do not appear to be influential. The goal is inference via model predictions.

Each model includes two random effect terms:
  - site-level: to account for the clustering of trees within survey sites;
  - species-level: to capture any systematic variation between species.


## Data

The data are stored as two non-spatial layers in the project geopackage. 

Tree-level observations...

```{r}

DAT_trees_density <- st_read(GPKG, "trees_density", quiet = TRUE)

str(DAT_trees_density)

```

Site-level predictors...

```{r}

DAT_predictors <- st_read(GPKG, "predictors_all", quiet = TRUE)

```

Sites...

```{r}

DAT_sites_density <- st_read(GPKG, "sites_density", quiet = TRUE)

```


Prepare a training data frame that includes all three broad bark types: smooth, stringy and rough.

```{r}

dat_all_barks_density <- DAT_trees_density %>%
  filter(for_analysis,
         !is.na(bark_density),
         bark_type_h %in% c("Smooth", "Stringy", "Rough", "Stocking")) %>%
  
  # Drop records for indet taxa (e.g. Eucalyptus 'unknown mallee')
  filter(!grepl(display_name, pattern = "unknown", ignore.case = TRUE)) %>%
  
  select(site_label, display_name, bark_type_h, bark_density, dbh) %>%

  mutate(f_species = factor(display_name)) %>%
  select(-display_name) %>%

  left_join(DAT_predictors, by = join_by(site_label)) %>%

  mutate(f_site = factor(site_label))

# Check data structure
str(dat_all_barks_density)

```


### Exploratory data summaries and graphs

Examine the number of survey sites in each combination of bark type and fire frequency.

```{r}

dat_all_barks_density %>%
  group_by(bark_type_h, fire_frequency) %>%
  summarize(nsites = n_distinct(f_site), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = bark_type_h, values_from = nsites, values_fill = 0)

```


Similar to the data preparation for the bark thickness models, we lump the fire frequency values into two classes: less frequent (<=3 fires recorded) versus more frequency (>=4 fires recorded).

```{r}

dat_all_barks_density <- dat_all_barks_density %>%
  mutate(f_firefreq = factor(ifelse(fire_frequency <= 3, "Less frequent fire", "More frequent fire")))

```


Summary table for the lumped fire frequency classes across bark types:

```{r}

dat_all_barks_density %>%
  group_by(bark_type_h, f_firefreq) %>%
  summarize(nsites = n_distinct(f_site), .groups = "drop") %>%
  tidyr::pivot_wider(names_from = bark_type_h, values_from = nsites, values_fill = 0)

```


## Model smooth bark density

```{r}

dat_smooth_bark_density <- dat_all_barks_density %>%
  filter(bark_type_h == "Smooth") %>%
  mutate(dbh_log = log(dbh))

```


Graph distribution of bark density values to check for any suspicious values.

```{r}

ggplot(data = dat_smooth_bark_density, aes(x = bark_density)) +
  geom_density() +
  geom_rug()

```


There is a value very close to zero...

```{r}

i <- which.min(dat_smooth_bark_density$bark_density)

dat_smooth_bark_density[i,]

```

Drop this value as potentially mistaken.

```{r}

dat_smooth_bark_density <- dat_smooth_bark_density[-i, ]

glue("{nrow(dat_smooth_bark_density)} observations for {n_distinct(dat_smooth_bark_density$site_label)} sites")

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_smooth_bark_density.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_smooth_bark_density <- readRDS(outpath)
  
} else {
  m_smooth_bark_density <- gam(
    bark_density ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_smooth_bark_density,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_smooth_bark_density, outpath)
}

summary(m_smooth_bark_density)
  
```


Posterior predictive check. Compare the distribution of bark density values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_smooth_bark_density, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_smooth_bark_density$bark_density,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_smooth_bark_density, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_density_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_smooth_bark_density, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_smooth_bark_density, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_density_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```


### Counter-factual predictions

Derive prediction with DBH held at its median value and the species and site random effects excluded.

```{r}

d_ <- dat_smooth_bark_density

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- dat_smooth_bark_density %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_smooth_bark_density, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_qi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_smooth_bark_density,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_density_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_smooth_bark_density

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_smooth_bark_density_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```


## Model stringy bark density

The model structure is based on that for bark thickness.

```{r}

dat_stringy_bark_density <- dat_all_barks_density %>%
  filter(bark_type_h == "Stringy") %>%
  mutate(dbh_log = log(dbh))

glue("{nrow(dat_stringy_bark_density)} observations for {n_distinct(dat_stringy_bark_density$site_label)} sites")

```


Graph distribution of bark density values to check for any suspicious values.

```{r}

ggplot(data = dat_stringy_bark_density, aes(x = bark_density)) +
  geom_density() +
  geom_rug()

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_stringy_bark_density.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_stringy_bark_density <- readRDS(outpath)
  
} else {
  m_stringy_bark_density <- gam(
    bark_density ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_stringy_bark_density,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_stringy_bark_density, outpath)
}

summary(m_stringy_bark_density)
  
```


Posterior predictive check. Compare the distribution of bark density values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_stringy_bark_density, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_stringy_bark_density$bark_density,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_stringy_bark_density, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_density_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_stringy_bark_density, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_stringy_bark_density, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_density_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```


### Counter-factual predictions

Derive prediction with DBH held at its median value and the species and site random effects excluded.

```{r}

d_ <- dat_stringy_bark_density

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- d_ %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_stringy_bark_density, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_qi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_stringy_bark_density,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_density_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_stringy_bark_density

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_stringy_bark_density_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```



## Model rough bark density

```{r}

dat_rough_bark_density <- dat_all_barks_density %>%
  filter(bark_type_h == "Rough") %>%
  mutate(dbh_log = log(dbh))

glue("{nrow(dat_rough_bark_density)} observations for {n_distinct(dat_rough_bark_density$site_label)} sites")

```


### Fit model

```{r}

# Path for fitted model
outpath <- file.path(MODEL_DIR, "gam_mgcv_rough_bark_density.rds")

if (!FORCE_REFIT && file.exists(outpath)) {
  m_rough_bark_density <- readRDS(outpath)
  
} else {
  m_rough_bark_density <- gam(
    bark_density ~ f_firefreq + 
      s(dbh_log) +
      s(aridity) +
      s(f_species, bs="re") +
      s(f_site, bs="re"),
    
    data = dat_rough_bark_density,
    #family = Gamma(link = "log"),
    family = tw(link = "log"),
    method = "REML")

  saveRDS(m_rough_bark_density, outpath)
}

summary(m_rough_bark_density)
  
```


Posterior predictive check. Compare the distribution of bark density values from the model (observation-level predictions) with the distribution of the observed data.

```{r}

post <- gratia::posterior_samples(m_rough_bark_density, n = 100, seed = SEED) %>%
  tidyr::pivot_wider(names_from = .row, values_from = .response) %>%
  select(-.draw) %>%
  as.matrix()

bayesplot::ppc_dens_overlay(y = dat_rough_bark_density$bark_density,
                            yrep = post)  

```


Graph the fitted smooth terms.

```{r}

gg <- gratia::draw(m_rough_bark_density, select = c("s(dbh_log)", "s(aridity)"))

outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_density_trends.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


### Species random effect

```{r}

### Random effect estimates with 50% and 90% bounds
x50 <- mixedup::extract_random_effects(m_rough_bark_density, re = "f_species", ci_level = 0.5)
x90 <- mixedup::extract_random_effects(m_rough_bark_density, re = "f_species", ci_level = 0.9)

# Extract estimates and format for ggplot, with species ordered by mean
# random effect value
dat_raneff <- x50 %>%
  bind_cols(x90 %>% select(lower_5, upper_95)) %>%
  rename(mean = value) %>%
  arrange(mean) %>%
  mutate(ordered_group = factor(group, levels = group, labels = group))


gg <- ggplot(data = dat_raneff, aes(y = ordered_group)) +
  # 90% intervals as thin lines
  geom_linerange(aes(xmin = lower_5, xmax = upper_95),
                 linewidth = 0.75, colour = "grey40") +

  # Mean points and 50% intervals as thicker lines  
  geom_pointrange(aes(xmin = lower_25, xmax = upper_75, x = mean),
                  linewidth = 1.5, size = 0.75, colour = "grey40") +
  
  labs(x = "Random effect estimate", y = "") +
  
  theme_bw()


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_density_spp_raneff.png")

ggsave(plot = gg, filename = outpath, width = 1800, height = 2400, units = "px")

print(gg)

```


### Counter-factual predictions

Derive prediction with DBH held at its median value and the species and site random effects excluded.

```{r}

d_ <- dat_rough_bark_density

dat_predict <- expand.grid(
  dbh_log = log(median(d_$dbh)),
  f_firefreq = factor(levels(d_$f_firefreq)),
  f_species = d_$f_species[1],
  f_site = d_$f_site[1],
  
  aridity = seq(min(d_$aridity), max(d_$aridity), length.out = 51)
)

# Trim the prediction data to the range of aridity seen within each fire frequency class
dat_arid_lims <- d_ %>%
  group_by(f_firefreq) %>%
  summarize(lo = min(aridity) - 0.02, hi = max(aridity) + 0.02)  # wiggle room

dat_predict <- dat_predict %>%
  left_join(dat_arid_lims, by = join_by(f_firefreq)) %>%
  filter(aridity >= lo & aridity <= hi)

NDRAWS <- 4000

# Draw observation-level predictions from the posterior distribution and summarize
# as 90% posterior prediction interval
dat_postpred <- gratia::add_posterior_samples(dat_predict, 
                                              model = m_rough_bark_density, 
                                              n = NDRAWS, 
                                              exclude = c("s(f_site)", "s(f_species)"),
                                              unconditional = TRUE,
                                              seed = SEED) %>%
  
  group_by(.row, aridity, f_firefreq) %>%
  
  tidybayes::median_qi(.response, .width = 0.9)


# Draw mean predictions from the posterior distribution and calculate 50% and 90% bounds
# (highest density intervals)
dat_epred <- gratia::add_fitted_samples(dat_predict,
                                        model = m_rough_bark_density,
                                        n = NDRAWS,
                                        seed = 99,
                                        scale = "response",
                                        exclude = c("s(f_site)", "s(f_species)"),
                                        unconditional = TRUE) %>%
  group_by(.row, aridity, f_firefreq)

dat_epred_mid <- dat_epred %>%
  summarize(mid = median(.fitted), .groups = "drop")

dat_hdi50 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.5) %>%
  select(.row, lwr50 = .lower, upr50 = .upper)

dat_hdi90 <- dat_epred %>%
  tidybayes::mean_hdi(.fitted, .width = 0.9) %>%
  select(.row, lwr90 = .lower, upr90 = .upper)

dat_gg <- dat_epred_mid %>%
  left_join(dat_hdi50, by = join_by(.row)) %>%
  left_join(dat_hdi90, by = join_by(.row))
  

gg <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_density_cfpredict.png")

ggsave(plot = gg, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg)

```


Alternative graph only colour observations for those species that occur near extremes of aridity.

```{r}

d_ <- dat_rough_bark_density

aridity_quantiles <- quantile(d_$aridity, c(0.1, 0.9))

# Identify species with occurrences at either aridity extreme
dat_focal_species <- d_ %>%
  group_by(f_species) %>%
  
  summarize(
    at_dry = any(aridity <= aridity_quantiles[1]),
    at_wet = any(aridity >= aridity_quantiles[2])
  ) %>%
  
  filter(at_dry | at_wet) %>%
  
  pull(f_species)


gg_focalspp <- ggplot(data = dat_gg, aes(x = aridity)) +
  # 50% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr50, ymax = upr50), alpha = 0.2) +
  
  # 90% bounds on predicted mean
  geom_ribbon(aes(ymin = lwr90, ymax = upr90), colour = "grey", alpha = 0.2) +
  
  # Mean trend
  geom_line(aes(y = mid), colour = "black", linewidth = 1, linetype = "solid", alpha = 0.8) +
  
  # 90% prediction interval (i.e. individual tree level)
  geom_ribbon(data = dat_postpred, aes(ymin = .lower, ymax = .upper),
              fill = NA, colour = "black", linetype = "longdash") +
  
  # Observed data
  geom_point(data = d_, aes(y = bark_density), colour = "grey", alpha = 0.7) +

  # Focal species observations coloured
  geom_point(data = d_ %>% filter(f_species %in% dat_focal_species),
             aes(y = bark_density, colour = f_species), alpha = 0.7) +
  
  labs(x = "Landscape aridity", y = "Bark density", colour = "Species") +
  
  facet_wrap(~f_firefreq) +
  
  scale_colour_viridis_d(option = "viridis") +
  
  theme(
    legend.text     = element_text(face = "italic", size = 8),
    legend.key.size = unit(0.3, "cm")
  )


outpath <- file.path(OUTPUT_DIR, "model_mgcv_rough_bark_density_cfpredict_focalspp.png")
ggsave(plot = gg_focalspp, filename = outpath, width = 2400, height = 1800, units = "px")

print(gg_focalspp)

```



